### 2016-06-29 store-consistency.tex
Edit.
edits to store-consistency

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicing the
behaviour of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

@@

A strongly-consistent {\sc sc} store guarantees a total order of all
operations w.r.t $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{sc}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both invariants is to rule out all 
executions that interleave the operations of one transaction with the
other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
inconsistent) with $\I_c$, the only way to enforce both invariant sets
is to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes many kinds of relational database-like stores.

For example, the semantics of a causally consistent data
store~\cite{gotsmanpopl16,LBC16} can also be obtained by defining
$\I_s$ to be the causal consistency ({\sc cc}) invariant that is
appropriately weakened to resolve conflicts with $\I_c$.\footnote{Details
of this instantiation are relegated to the supplementary in the interest of space.}

@@

Reasoning under weakly consistent replication has received special
attention in recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case. In this section, we explain how the standard artifacts
of weakly-consistent replication manifest in our reasoning framework.

Reasoning about weakly-consistent replication involves reasoning in
terms of \emph{replicas} of data. The primary challenge in this
setting is to ensure that the assumptions made and guarantees enforced
by an operation at one replica carry over to other replicas that merge
their effects, thus preserving the overall integrity of the system.
In~\cite{gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to stores (e.g., \cite{bayou,pldi15}) that provide
guarantees weaker than causal consistency, such as causal visibility
or read-my-writes~\cite{zoo}. Causal consistency is not baseline
consistency in these stores because it is not \emph{highly
available}~\cite{bailishat}.

Notably, our view of replication however does not explicitly involve
replicas.  Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based
model.  Under our model, the notion of a replica is subsumed by the
concept of visibility; a replica is defined by the subset ($S$) of
global state ($\E.\A$) that an operation witnesses. Constraints over
replica states therefore manifest as constraints over a specific visibility
relation.  For example, instead of requiring the store to be causally
consistent, an operation can request to witness a causally consistent
subset of the state; such demands can be made via the trace invariant
$\I$. For a precondition ($P$) of the operation to be useful, it has
to be an assertion over every causally consistent subset of the global
state.  Since any replica that eventually executes the operation has
to expose one such subset ($S$), the precondition is guaranteed to
hold regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for this
change,~\cite{gotsmanpopl16} distinguishes between effect generation
event at one replica $r_1$ and effect merge event at replica $r_2$,
and requires that only \emph{non-conflicting} operations execute
between these two events at $r_2$, and that they preserve certain
invariants. Our framework folds all such machinery into a stability
condition predicated on $\I$ (\S\ref{sec:rely-guarantee}).  Since any
change to the global state during the execution of the operation is an
interference, and $P$ is required to be stable with respect to any
such interference, it follows that $P$ is valid on every replica, thus
ensuring that assumptions made at the generation event are also valid
at the merge event.

@@

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. The proof for
each triple follows from the rule for the corresponding term in
Fig.~\ref{fig:rg-rules}. For illustration, we consider the triple
corresponding to the assignment statement (line ??). The precondition
($\phi_4$) asserts that \C{Wd2} writes to \C{C}, and the value of the
temporary variable \C{v3} is \C{k-a2-a1}. The RG rule
(\rulelabel{RG-Asgn}) for assignment statement allows us to easiliy
conclude the following about the execution state after the assignment:
\begin{smathpar}
\begin{array}{l}
    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k}) \conj\\
    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
    \C{C}) \conj \\
    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
\end{array}
\end{smathpar}
The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions write to \C{C}, {\sc si} requires either $\C{Wd1}
\visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1} is not yet
committed ($\neg\committed(\C{Wd1})$), {\sc si} prohibits the former
possibility, allowing us the deduce the $\C{Wd2} \visar \C{Wd1}$. This
lets us derive ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge
\C{C = k-a2-a1}$, allowing us to prove the postcondition ($\phi_5$).

>>>>

The operational semantics of Fig.~\ref{fig:txnimp} allows operations
to witness arbitrary subsets of the global state, effectively mimicking the
behaviour of an eventually consistent ({\sc ec}) data
store\footnote{Eventual consistency guarantees that in the absence of further
  updates, all reads witness the same global state
  \emph{eventually}. In any finite trace, however, there are no
  guarantees on what a read may witness.}. There are, however, data stores,
such as relational databases, that provide stronger consistency
guarantees than {\sc ec}. Like the isolation levels of transactions,
the consistency level of the underlying store also affects the
semantics of a program in non-trivial ways. In this section, we
demonstrate how the semantics of stronger stores with on-demand weak
isolation can be captured in our operational model.

@@

A strongly-consistent {\sc sc} store guarantees a total order on all
operations w.r.t $\visZ$ consistent with their chronological order. A
straightforward $\I_s$ for this store is the \C{SC} property
formalized below:
\begin{smathpar}
\begin{array}{l}
  \C{sc}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$, a constraint that dictates all
transactions execute under \iso{Read Committed} isolation. Imagine a
sample execution where $\eta_1$'s transaction is not yet committed
when $\eta_2$ is generated. Letting $\eta_1$ be visible to $\eta_2$
violates $\I_c$, whereas not letting it be visible violates $\I_{s}$.
The only way to satisfy both invariants is to rule out all 
executions that interleave the operations of one transaction with the
other, thereby enforcing serializability.  In general, when $\I_s$ conflicts (but is not
inconsistent) with $\I_c$, the only way to enforce both invariant sets
is to restrict concurrency. Clearly, this is unacceptable since it
defeats the very purpose of supporting weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

In practice, relational databases resolve such conflicts by
prioritizing weak isolation (thus, concurrency and performance) over
strong consistency, so the execution traces do not necessarily satisfy
{\sc sc}. In particular, visibility constraints imposed by {\sc sc}
are violated iff they are found to be in conflict with the constraints
imposed by a transaction's isolation level. In the context of
the aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
    commitment}~\cite{bailishat} is often used in practice
  to capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
  visiblity principle} to determine an acceptable weakening of $\I_s$
in case of a conflict with $\I_c$.  The principle requires the
weakened consistency guarantee ($\I_s'$) of the store to enforce all
visibility relationships imposed by the actual consistency guarantee
($\I_s$), unless enforcing such a relationship violates $\I_c$.
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
    {\sc sc} except in cases where they are in conflict with $\I_c$.
    Instantiating the parameter $\I$ with $\I_s \wedge \I_c$ in
    Fig.~\ref{fig:txnimp} results in an operational semantics that
    describes many kinds of relational database-like data stores,
    including causally-consistent data stores, among
    others.\footnote{Details of this instantiation are relegated to
      the supplementary in the interest of space.}

@@

Reasoning under weakly-consistent replication has received special
attention in recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case of our formulation. In this section, we explain how the
standard artifacts of weakly-consistent replication manifest in our
reasoning framework.

The primary challenge in this setting is to ensure that the
assumptions made and guarantees enforced by an operation at one
replica carry over to other replicas that merge their effects, thus
preserving the overall integrity of the system.  In prior
work~\cite{lbc16,gotsmanpopl16}, this challenge is partly addressed by
imposing restrictions on how various replica states differ, i.e., by
fixing a system model with a stronger baseline consistency ({\sc cc})
than {\sc ec}. This unfortunately restricts the reasoning approach
from being applied to data stores (e.g., \cite{bayou,pldi15}) that provide
guarantees weaker than causal consistency, such as causal visibility
or read-my-writes~\cite{zoo}. Causal consistency is not baseline
consistency in these stores because it is not \emph{highly
  available}~\cite{bailishat}.

Notably, our view of replication does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over a specific visibility relation.
For example, instead of requiring the store to be causally consistent,
an operation can witness a causally consistent subset of the state;
such demands can be made via the trace invariant $\I$. For a
precondition ($P$) of the operation to be useful, it has to be an
assertion over every causally consistent subset of the global state.
Since any replica that eventually executes the operation has to expose
one such subset ($S$), the precondition is guaranteed to hold
regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for such
changes, we might choose to distinguish between effect generation
event at one replica $r_1$ and effect merge event at replica $r_2$,
requiring that \emph{non-conflicting} operations execute between these
two events at $r_2$, and that they preserve certain
invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
machinery into a stability condition predicated on $\I$
(\S\ref{sec:rely-guarantee}).  Since any change to the global state
during the execution of the operation is an interference, and $P$ is
required to be stable with respect to any such interference, it
follows that $P$ is valid on every replica, thus ensuring that
assumptions made at a generation event is also valid at the merge
event.

@@

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. The proof for
each triple follows from the rule for the corresponding term in
Fig.~\ref{fig:rg-rules}. For illustration, we consider the triple
corresponding to the assignment statement (line ??). The precondition
($\phi_4$) asserts that \C{Wd2} writes to \C{C}, and the value of the
temporary variable \C{v3} is \C{k-a2-a1}. The RG rule
(\rulelabel{RG-Asgn}) for assignment statement allows us to easiliy
conclude the following about the execution state after the assignment:

\begin{smathpar}
\begin{array}{l}
    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k}) \conj\\
    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
    \C{C}) \conj \\
    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
\end{array}
\end{smathpar}

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions write to \C{C}, {\sc si} requires either $\C{Wd1}
\visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1} is not yet
committed ($\neg\committed(\C{Wd1})$), {\sc si} prohibits the former
possibility, allowing us the deduce the $\C{Wd2} \visar \C{Wd1}$. This
lets us derive ${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge
\C{C = k-a2-a1}$, allowing us to prove the postcondition ($\phi_5$).
